Velocity Map Imaging Spectroscopy of C$_2$H$^-$ and C$_2$D$^-$: a benchmark study of vibronic coupling interactions

### 2022-05-05 C2Hpaper.tex
include figs

\section{Introduction}
Carbon monohydrides C$_{2n}$H, are a class of linear radicals that play an important role in combustion and interstellar chemistry~\cite{fre85,bou96,wil91,vui01}. These carbon chains have been observed in many interstellar environments, including planetary atmospheres~\cite{vui01,wil03,dob16}, comets~\cite{jac96}, dark clouds~\cite{ziu82,gup09}, and during heavy star formation~\cite{beu08,jia15}. Due to their relatively high abundance in a variety of astronomical conditions, they are believed to be promising candidates for Diffuse Interstellar Band (DIB) carriers~\cite{dou77,ful93,wat94,ful00,sch05,gup09}. The corresponding anions C$_{2n}$H$^-$ are also believed to play an important role in interstellar chemistry~\cite{mil17,gup09}. C$_6$H$^-$ was the first negative ion detected in space, after it was observed in two distinct interstellar molecular clouds IRC+10216 and TMC-1 in 2006~\cite{mcc06}. More recently, anions C$_4$H$^-$~\cite{cer07} and C$_8$H$^-$~\cite{bru07,rem07} have also been detected in a range of environments, include dark clouds~\cite{cor13}, prestellar cores,~\cite{sak10} and prostellar envelopes~\cite{sak07}.

@@

These considerations may be explored by examining some of the smallest carbon monohydrides. The ethynyl radical C$_2$H may appear to be a simple linear triatomic molecule. However the electronic spectrum is complicated by the presence of the close-lying ground $^2\Sigma^+$ and first excited $^2\Pi$ surfaces, which are only separated by $\sim3,700~$cm$^{-1}$~\cite{cur85,tar03}. The interaction of these surfaces produces a complex vibronic spectrum around the $\tilde{A}^2\Pi$ state, where the $\tilde{X}-\tilde{A}$ origin is spread over multiple admixed vibronic levels. In C$_4$H the $^2\Sigma^+$ and $^2\Pi$ states are nearly degenerate, resulting in even stronger coupling~\cite{zho07}, while in C$_6$H and C$_8$H the ordering of the states swaps, with a ground $^2\Pi$ state and a low lying excited $^2\Sigma^+$ state~\cite{lin99,tay98}. Consequently, understanding the vibronic coupling interactions between the $^2\Sigma$ and $^2\Pi$ surfaces is essential in order to accurately model the role these radicals (and their corresponding anions) are likely to play in the interstellar chemistry discussed above, and to guide the search for possible DIB carriers.

@@

These interactions spread the electronic orgin of the $\tilde{A}\,^2\Pi$ state over several vibronic levels. Therefore, instead of assigning a defined origin, the observed peaks in the spectrum may be assigned to coupled admixtures of vibronic transitions involving the three potential energy surfaces,
\begin{equation}
\Psi_f = \sum_\xi \psi_e^\xi \sum_k C_{fk}^\xi\phi_{fkm}^\xi,
\label{eq:teller3} 
\end{equation}
where $\psi_e^\xi$ is the diabatic electronic wavefunction, and $\phi_{fkm}^\xi$ is the spin-rovibrational wavefunction. $\xi$ represents the electronic states used in the expansion ($\xi=\Sigma^+(1A'),~\Pi^+(2A'),~\Pi^-(1A'')$). A depiction of these three interacting surfaces, generated from the parameters in Ref.~[\onlinecite{tar03}], is given in Fig.~\ref{fig:2}.

@@

\begin{equation}
\eta = \frac{1}{2}\left[ f_{2a2a}^{A(a')}-f_{2b2b}^{A(a')}-\frac{2\lambda^2}{(\Delta_0^A-\Delta_0^X)}\right]
\label{eq:calc3}
\end{equation}

@@

\begin{table}
	\caption{Parameters for the quasidiabatic Hamiltonian Eq.~(\ref{eq:calc1}) determined using CFOUR and the quasidiabatic ansatz. Units are in cm$^{-1}$.}
	\label{tab:1}
	\begin{tabular}{c |c c c c| c c c c }
		 & \multicolumn{4}{c}{C$_2$H}     & \multicolumn{4}{c}{C$_2$D}  \\ \hline
		  & Anion & $\tilde{X}^2\Sigma^+$ & $\tilde{A}^2\Pi (a')$ & $\tilde{A}^2\Pi (a'')$ & Anion & $\tilde{X}^2\Sigma^+$ & $\tilde{A}^2\Pi (a')$ & $\tilde{A}^2\Pi (a'')$ \\
		 \hline\hline
		 $\Delta$E & 0 & 22431.21 &  26543.67 & 26543.67 & 0&	22431.21 & 26543.67	& 26543.67\\
		 f1 & 0	& -211.28 & 286.9 &	286.9 & 0& -596.04& 641.6 & 641.6\\
		 f2 & 0 & -1515.21 & 1257.07 & 1257.07 & 0 & -1401.53 & 1115.17 & 1115.17\\
		 f3 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
		 f4 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
         f11 & 3424.32 & 3487.03 & 3482.16 & 3482.16 & 2611.42 & 2636.57 & 2681.25 & 2681.25\\
         f12 & 0.00 & -28.32 & 31.46 & 31.46 & 0.00 & -48.38 & 55.40 & 55.40\\
         f22 & 1894.71 & 1830.83 & 2099.57 & 2042.21 & 1792.49 & 1747.62 & 1914.39 & 1914.39\\
         f33 & 607.09 & 87.57 & 1883.48 & 658.96 & 481.89 & 120.09 & 1503.88 & 473.55\\
         f44 & 607.09 & 87.57 & 658.96 & 1883.48 & 481.89 & 120.09 & 473.55 & 1503.88\\
         $\lambda$ &  & & -1185.26 &  &  & & -1014.58 &  \\
         \end{tabular}
\end{table}

@@

The results from Figures~\ref{fig:3} and~\ref{fig:4} confirm that the quasidiabatic approach is able to accurately describe the vibronic interactions between the $^2\Sigma^+$ and $^2\Pi$ surfaces, including deuteration effects. This validates the proposed vibronic interactions, and demonstrates how this method can be employed to decode and assign even complex spectra. While the present calculations invoke a high level of \emph{ab-initio} theory (EOM-CCSDT), the level of parametrization of the vibronic Hamiltonian is relatively simple (QVC model). The approach could be systematically improved with a more elaborate expansion, such as that used in Ref.~[\onlinecite{sim12}] for the NO$_3$ radical. Applying this approach to similar less-studied systems will produce reliable predictions for the position and intensity of dominant transitions, which will help guide the search for these molecules in laboratory experiments and astronomical observations.

@@

The sign (+/-) of $\beta$ can be used to assign each accessed vibronic level to either the $^2\Sigma^+$ or $^2\Pi$ vibronic symmetry respectively. This is particularly useful around the $\tilde{A} ^2\Pi$ origin, where the two states overlap. It is important to note that the HT coupled transitions on the $\tilde{X} ^2\Sigma^+$ surface have negative anisotropies and $\Pi$ symmetry. This is discussed in more detail in the Supplementary Materials.

A plot of the anisotropy parameters for C$_2$H$^-$ detachment at 300~nm, represented as $\beta\times I$, is presented in Figure~\ref{fig:6}, alongside the experimental spectrum. Plotting $\beta\times I$ allows the sign (+/-) of each individual transition to be easily identified, even for partially resolved peaks. All of the transitions above the $\tilde{A} ^2\Pi$ origin have a negative anisotropy, except for peak $f$. This confirms the assignment of this peak to the highly excited $\Sigma$ transition $\tilde{X}(0,6,1$), where $\tilde{X}(v_1,v_2,v_3)$. The nearby peaks $h$ and $i$  are assigned to the admixed $\Pi$ transitions $\tilde{X}(0,3,2)\tilde{A}(0,0,0)$ and $\tilde{X}(0,7,1)\tilde{A}(010)$. The other two peaks in this region $j$ and $k$ are assigned to highly excited HT coupled $\Pi$ transitions $\tilde{X}(0,7,1)$ and $\tilde{X}(0,11,0)$. The assignments for all peaks in the C$_2$H$^-$ and C$_2$D$^-$ photoelectron spectra are presented in the Supplementary Materials Table~S1.

@@

\section{Conclusions} 
High-resolution photoelectron images of C$_2$H$^-$ and C$_2$D$^-$ were recorded both near and above threshold, to investigate vibronic coupling effects in the simplest C$_{2n}$H carbon monohydride chain. The interplay of pseudo Jahn-Teller and Renner Teller coupling between the close lying $\tilde{X}^2\Sigma^+$ and $\tilde{A}^2\Pi$ states results in a complex vibronic spectrum near the $\tilde{A}$ state surface, with the electronic origin split over multiple admixed levels.We have shown that by transforming to a quasidiabatic representation, the effect of these coupling interactions may be accurately simulated. A KDC Hamiltonian was constructed, and parametrised using the high-level EOMIP-CCSDT quantum chemical approach and the quasidiabatic ansatz. The resulting simulated photoelectron spectra of C$_2$H$^-$ and C$_2$D$^-$ closely resemble the experimental data, even in the energy region of the complicated $\tilde{A}$ state origin. Photoelectron anisotropy parameters were also measured for each dominant transition, at a range of wavelengths. Comparing the sign (+/-) of the anisotropy to the calculated vibronic symmetry allowed for even weak or partially resolved transitions to be assigned.

The excellent agreement between the experimental and calculated spectra in this work
demonstrates that electronic spectra complicated by vibronic coupling effects may still be accurately simulated using a quasidiabatic treatment, as parametrised by appropriate quantum chemical methodology. This approach may be applied to larger, less well known, C$_{2n}$H chains, which may help guide the search for these molecules in astronomical studies, and help identify potential DIB carriers. It is especially encouraging that the fairly modest parametrisation used here works so well for C$_2$H, as this bodes well for its extension to larger systems.

@@

A modified velocity-map imaging lens images the detached photoelectrons to a 75mm diameter MCP/phosphor screen detector. Events are imaged by a 2048x2048 monochrome CCD camera (PCO 2000), with each frame transferred to a computer at a 10Hz repetition rate, and processed in real time to identify individual electron events. The electron positions are centroided to a sub-pixel accuracy, then written to a data file for subsequent analysis. The velocity-map image is centred and then circularized by an angular dependent-radial scaling determined by comparing adjacent radial slice intensity profiles~\cite{gas17}. An inverse Abel transformation of the VMI, based on the algorithm of Hansen and Law~\cite{han85,hic19}, returns a slice image of the 3D electron source distribution. Absolute energy calibration of the photoelectron spectra is achieved using published measurements of species, including O$^-$~\cite{cav07} and NO$_2^-$~\cite{law19}, that have been studied under similar conditions as used for the C$_2$H$^-$ measurements.

>>>>

\section{Introduction}
Carbon monohydrides C$_{2n}$H, are a class of linear radicals that play an important role in combustion and interstellar chemistry~\cite{kie90,kie92,bou96,wil91,vui01}. These carbon chains have been observed in many interstellar environments, including planetary atmospheres~\cite{vui01,wil03,dob16}, comets~\cite{jac96}, dark clouds~\cite{ziu82,gup09}, and during heavy star formation~\cite{beu08,jia15}. Due to their relatively high abundance in a variety of astronomical conditions, they are believed to be promising candidates for Diffuse Interstellar Band (DIB) carriers~\cite{dou77,ful93,wat94,ful00,sch05,gup09}. The corresponding anions C$_{2n}$H$^-$ are also believed to play an important role in interstellar chemistry~\cite{mil17,gup09}. C$_6$H$^-$ was the first negative ion detected in space, after it was observed in two distinct interstellar molecular clouds IRC+10216 and TMC-1 in 2006~\cite{mcc06}. More recently, anions C$_4$H$^-$~\cite{cer07} and C$_8$H$^-$~\cite{bru07,rem07} have also been detected in a range of environments, include dark clouds~\cite{cor13}, prestellar cores,~\cite{sak10} and prostellar envelopes~\cite{sak07}.

@@

These considerations may be explored by examining some of the smallest carbon monohydrides. The ethynyl radical C$_2$H may appear to be a simple linear triatomic molecule. However the electronic spectrum is complicated by the presence of the close-lying ground $^2\Sigma^+$ and first excited $^2\Pi$ surfaces, which are only separated by $\sim3,700~$cm$^{-1}$~\cite{cur85,tar03}. The interaction of these surfaces produces a complex vibronic spectrum around the $\tilde{A}^2\Pi$ state, where the $\tilde{X}-\tilde{A}$ origin is spread over multiple admixed vibronic levels. In C$_4$H the $^2\Sigma^+$ and $^2\Pi$ states are nearly degenerate, resulting in even stronger coupling~\cite{zho07}, while in C$_6$H and C$_8$H the ordering of the states swaps, with a ground $^2\Pi$ state and a low lying excited $^2\Sigma^+$ state~\cite{lin99,tay98}. Consequently, understanding the vibronic coupling interactions between the $^2\Sigma^+$ and $^2\Pi$ surfaces is essential in order to accurately model the role these radicals (and their corresponding anions) are likely to play in the interstellar chemistry discussed above, and to guide the search for possible DIB carriers.

@@

These interactions spread the electronic orgin of the $\tilde{A}\,^2\Pi$ state over several vibronic levels. Therefore, instead of assigning a defined origin, the observed peaks in the spectrum may be assigned to coupled admixtures of vibronic transitions involving the three potential energy surfaces,
\begin{equation}
\Psi_f = \sum_\xi \psi_e^\xi \sum_k C_{fk}^\xi\phi_{fkm}^\xi,
\label{eq:teller3} 
\end{equation}
where $\psi_e^\xi$ is the diabatic electronic wavefunction, and $\phi_{fkm}^\xi$ is the spin-rovibrational wavefunction. $\xi$ represents the electronic states used in the expansion ($\xi=\Sigma^+(1A'),~\Pi^+(2A'),~\Pi^-(1A'')$). A depiction of these three interacting surfaces, generated from the parameters in Ref.~\onlinecite{tar03}, is given in Fig.~\ref{fig:2}.

@@

\begin{equation}
\eta = \frac{1}{2}\left[ f_{2a2a}^{A(a')}-f_{2b2b}^{A(a'')}-\frac{2\lambda^2}{(\Delta_0^A-\Delta_0^X)}\right]
\label{eq:calc3}
\end{equation}

@@

\begin{table}
	\caption{Parameters for the quasidiabatic Hamiltonian Eq.~(\ref{eq:calc1}) determined using CFOUR and the quasidiabatic ansatz. Units are in cm$^{-1}$.}
	\label{tab:1}
	\begin{tabular}{c |c c c c| c c c c }
		 & \multicolumn{4}{c}{C$_2$H}     & \multicolumn{4}{c}{C$_2$D}  \\ \hline
		  & Anion & $\tilde{X}^2\Sigma^+$ & $\tilde{A}^2\Pi (a')$ & $\tilde{A}^2\Pi (a'')$ & Anion & $\tilde{X}^2\Sigma^+$ & $\tilde{A}^2\Pi (a')$ & $\tilde{A}^2\Pi (a'')$ \\
		 \hline\hline
		 $\Delta$E & 0 & 22431.21 &  26543.67 & 26543.67 & 0&	22431.21 & 26543.67	& 26543.67\\
		 F$_1$ & 0	& -211.28 & 286.9 &	286.9 & 0& -596.04& 641.6 & 641.6\\
		 F$_3$ & 0 & -1515.21 & 1257.07 & 1257.07 & 0 & -1401.53 & 1115.17 & 1115.17\\
		 F$_{2a}$ & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
		 F$_{2b}$ & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
         F$_{11}$ & 3424.32 & 3487.03 & 3482.16 & 3482.16 & 2611.42 & 2636.57 & 2681.25 & 2681.25\\
         F$_{13}$ & 0.00 & -28.32 & 31.46 & 31.46 & 0.00 & -48.38 & 55.40 & 55.40\\
         F$_{33}$ & 1894.71 & 1830.83 & 2099.57 & 2042.21 & 1792.49 & 1747.62 & 1914.39 & 1914.39\\
         F$_{2a2a}$ & 607.09 & 87.57 & 1883.48 & 658.96 & 481.89 & 120.09 & 1503.88 & 473.55\\
         F$_{2b2b}$ & 607.09 & 87.57 & 658.96 & 1883.48 & 481.89 & 120.09 & 473.55 & 1503.88\\
         $\lambda$ &  & & -1185.26 &  &  & & -1014.58 &  \\
         \end{tabular}
\end{table}

@@

The results from Figures~\ref{fig:3} and~\ref{fig:4} confirm that the quasidiabatic approach is able to accurately describe the vibronic interactions between the $^2\Sigma^+$ and $^2\Pi$ surfaces, including deuteration effects. This validates the proposed vibronic interactions, and demonstrates how this method can be employed to decode and assign even complex spectra. While the present calculations invoke a high level of \emph{ab-initio} theory (EOM-CCSDT), the level of parametrization of the vibronic Hamiltonian is relatively simple (QVC model). The approach could be systematically improved with a more elaborate expansion, such as that used in Ref.~\onlinecite{sim12} for the NO$_3$ radical. Applying this approach to similar less-studied systems will produce reliable predictions for the position and intensity of dominant transitions, which will help guide the search for these molecules in laboratory experiments and astronomical observations.

@@

The sign (+/-) of $\beta$ can be used to assign each accessed vibronic level to either the $^2\Sigma^+$ or $^2\Pi$ vibronic symmetry respectively. This is particularly useful around the $\tilde{A} ^2\Pi$ origin, where the two states overlap. It is important to note that the HT coupled transitions on the $\tilde{X} ^2\Sigma^+$ surface have negative anisotropies and $\Pi$ symmetry.

A plot of the anisotropy parameters for C$_2$H$^-$ detachment at 300~nm, represented as $\beta\times I$, is presented in Figure~\ref{fig:6}, alongside the experimental spectrum. Plotting $\beta\times I$ allows the sign (+/-) of each individual transition to be easily identified, even for partially resolved peaks. All of the transitions above the $\tilde{A} ^2\Pi$ origin have a negative anisotropy, except for peak $f$, which was not assigned in previous experiments. The positive anisotropy supports our assignment of this peak to the highly excited $\Sigma$ transition $\tilde{X}(0,6,1$), where $\tilde{X}(v_1,v_2,v_3)$. The nearby peaks $h$ and $i$  are assigned to the admixed $\Pi$ transitions $\tilde{X}(0,3,2)\tilde{A}(0,0,0)$ and $\tilde{X}(0,7,1)\tilde{A}(010)$, which agrees with previous assignments~\cite{zho07,tar03}. Peak $j$ was tentatively assigned to the $\Sigma$ transition $\tilde{X}(1,4,0$) previously~\cite{zho07}, however Figure~\ref{fig:6} shows peak $j$ has $\Pi$ symmetry. Therefore, we assign $j$ to the highly excited HT coupled $\Pi$ transition $\tilde{X}(0,7,1)$. The other dominant peaks in this region, $l$, $n$, and $o$, represent admixed $\Pi$ transitions involving the $\tilde{A}(0,0,1)$ vibrational level of the excited state. The complete assignments for all resolved peaks in the C$_2$H$^-$ and C$_2$D$^-$ photoelectron spectra from this work are presented in the Supplementary Materials Table~S1.

@@

\section{Conclusions} 
High-resolution photoelectron images of C$_2$H$^-$ and C$_2$D$^-$ were recorded both near and above threshold, to investigate vibronic coupling effects in the simplest C$_{2n}$H carbon monohydride chain. The interplay of pseudo Jahn-Teller and Renner Teller coupling between the close lying $\tilde{X}^2\Sigma^+$ and $\tilde{A}^2\Pi$ states results in a complex vibronic spectrum near the $\tilde{A}$ state surface, with the electronic origin split over multiple admixed levels.We have shown that by transforming to a quasidiabatic representation, the effect of these coupling interactions may be accurately simulated. A KDC Hamiltonian was constructed, and parametrised using the high-level EOMIP-CCSDT quantum chemical approach and the quasidiabatic ansatz. The resulting simulated photoelectron spectra of C$_2$H$^-$ and C$_2$D$^-$ closely resemble the experimental data, even in the energy region of the complicated $\tilde{A}$ state origin. Photoelectron anisotropy parameters were also measured for dominant transitions, at a range of wavelengths. Comparing the sign (+/-) of the anisotropy to the calculated vibronic symmetry allowed for even weak or partially resolved transitions to be assigned.

The excellent agreement between the experimental and calculated spectra in this work
demonstrates that electronic spectra complicated by vibronic coupling effects may still be accurately simulated using a quasidiabatic treatment, as parametrised by appropriate quantum chemical methodology. This approach may be applied to larger, less well known, C$_{2n}$H species, which may help guide the search for these molecules in astronomical studies, and help identify potential DIB carriers. It is especially encouraging that the fairly modest parametrisation used here works so well for C$_2$H, as this bodes well for its extension to larger systems.

@@

A modified velocity-map imaging lens images the detached photoelectrons to a 75mm diameter MCP/phosphor screen detector. Events are imaged by a 2048x2048 monochrome CCD camera (PCO 2000), with each frame transferred to a computer at a 10Hz repetition rate, and processed in real time to identify the electron events, centroiding the position to
a subpixel accuracy. The electron positions are then written to a data file for subsequent analysis. The velocity-map image is centred and then circularized by an angular dependent-radial scaling determined by comparing adjacent radial slice intensity profiles~\cite{gas17}. An inverse Abel transformation of the VMI, based on the algorithm of Hansen and Law~\cite{han85,hic19}, returns a slice image of the 3D electron source distribution. Absolute energy calibration of the photoelectron spectra is achieved using published measurements of species, including C$_2^-$~\cite{law19b} and NO$_2^-$~\cite{law19}, that have been studied under similar conditions and photon energies as those used for the C$_2$H$^-$ measurements.
